---
title: "Course Introduction"
author: "Georgios Gousios"
date: "`r format(Sys.time(), '%d %B %Y')`"
nocite:
csl: ieee.csl
bibliography: bibliography.bib
---

```{r child="header.Rmd", include=FALSE}
```

## In this course, we will learn

* Programming techniques, specifically functional programming, useful
for Big Data Systems.
* Basic distributed systems theory, applied on databases and file systems
* How to process big and fast data with modern systems

## Tools and technologies

During our course we will learn how to process data with

* Spark, and related libraries
* Flink

We will learn to store out data in HDFS and MongoDB.

The course will be using the following 2 languages

* Scala: the de-facto standard for high performance data processing
* Python: the richest data analysis environment

You can submit your assignments in any of those!

## Grades

* 60% final exam, written + multiple-choice. Minimum grade 6.
* 40% assignments
* 10% optional assignment

There will be resit, there is no mid-term.

You can transfer your assignment grade AS A WHOLE. No individual assignment
resubmissions!

## Assignments

* The course consists of 4 obligatory assignments, and 1 optional one.

* You always work in groups

* We will be using [Jupyter notebooks](http://jupyter.org) to submit
the assignments.

* Grade: $\frac{\sum_{assign=1}^{5} grade(assign)}{5}$, aka the mean
    * No submission? Grade $\rightarrow$ 0
    * Late submission? Grade $\rightarrow$ 0
    * Non-show for sign-off? Grade $\rightarrow$ 0

* Most assignment grades add to $> 10$

## Assignment sign-off

Deadlines for each assignment are on the course's web site

**One day before the deadline** (before 23:59) you must:

* Upload a PDF version of your Jupyter notebook to Brighspace, at
a designated location
* Book a designated timeslot with an examiner, using a Doodle link

**On the deadline date**

* You show up at the designated timeslot
    * late arrivals are graded with 0
* You present a LIVE version of the Jupyter notebook
    * The examiner might ask you to re-run parts of the code

## Forming groups

All assignments are done in pairs

* The groups are formed semi-randomly per assignment.
    * You will be notified about your group
* Problems within the group? Talk to the TAs!

## Getting help

* Generally, two types of Wednesdays
  * Practicum Wednesdays: you work on you assignments and TAs provide help
  * Sign-off Wednesdays: you only need to be there for sign-off, but you can
  be there longer before/after to work on other assignments

* The TAs will add everybody to a Slack channel

## Course resources

* The course VM: Contains all software (Spark, HDFS, Flink) you may need pre-installed
    * Provided only for convenience: no technical support

* The course cluster
    * `bdp1.ewi.tudelft.nl` from within TU Delft or VPN.
    * Ask me for the password :-)
    * Dataset to upload that you would like to analyze? Email me the link!
    * The first rule of the BDP cluster: respect the others!

## Slide symbols

* **Q** A question with a known answer; this will be revealed, but we should
work together towards it!

* **D** A open discussion item; we need to think and discuss.

## Lecture notes

* Freely available on the web, on my homepage (http://gousios.org/teaching.html)

* You can print/download them before the lecture and bring them along to make
additional notes.

* I am looking forward to improve them! If you have suggestions etc, let me know!

```{r child="footer.Rmd", include=FALSE}
```
