---
title: "Course Introduction"
author: "Georgios Gousios"
date: "`r format(Sys.time(), '%d %B %Y')`"
nocite:
csl: ieee.csl
bibliography: bibliography.bib
---

```{r child="header.Rmd", include=FALSE}
```

## Algorithms and Data Structures

> **Algorithm**: Any well-defined computational procedure that takes some value, or set of values, as input and produces some value, or set of values, as output.

. . .

> **Data structure**: A particular way of storing and organising data in memory so that it can be used efficiently.

. . .

Programming is the _means_ to express a solution to a problem.

Algorithms and Data Structures are the heart of CS.

## Learning objectives

* Understand, explain, and implement standard data structures.
* Understand, explain, and implement standard algorithms.
* Apply standard data structures and algorithms to solve programming tasks.
* Analyse and compare implementations with respect to their time and space complexity.
* Understand and explain advanced topics in algorithm design

## Tools and technologies

* We will be using Python throughout the course.

* You can use your favourite editor or IDE (I recommend PyCharm) 

* We will be using [Jupyter notebooks](http://jupyter.org) to submit
the assignments.

* We will be using Brightspace for assignment uploads and sending announcements.
Remember to subscribe!

## Assignments

* The course work consists of 4 obligatory assignments.

* You always work in groups

* The groups are randomly created per assignment

* Grade: $\frac{\sum_{assign=1}^{4} grade(assign)}{4}$, aka the mean
    * No submission? Grade $\rightarrow$ 0
    * Late submission? Grade $\rightarrow$ 0
    * Non-show for sign-off? Grade $\rightarrow$ 0

* Most assignment grades add to $> 10$

## Grades

* 60% final exam, written + multiple-choice. Minimum grade 6.
* 40% assignments

There will be a resit, there is no mid-term.

You can transfer your assignment grade to the resit AS A WHOLE!

## Assignment sign-off

Deadlines for each assignment are on the course's web site

**One day before the deadline** (before 23:59) you must:

* Upload a PDF version of your Jupyter notebook to Brighspace, at
a designated location
* Book a designated timeslot with an examiner, using a Doodle link provided
to you

**On the deadline date**

* You show up at the designated timeslot
    * late arrivals are graded with 0
* You present a LIVE version of the Jupyter notebook
    * The examiner might ask you to re-run parts of the code

## Forming groups

All assignments are done in pairs:

* You can decide your partner; fill in the Google form the TAs will distribute.
* No partner? Ask the TAs!
* Problems within the group? Talk to the TAs!

## Getting help

* Generally, two types of Thursdays
  * Practicum Thursdays: you work on you assignments and TAs provide help
  * Sign-off Thursdays: you only need to be there for sign-off, but you can
  be there longer before/after to work on other assignments

* The TAs will add everybody to a Slack channel

## TAs and their role

The TAs are there to help you! They are responsible for

* facilitating your learning
* grading your assignments and giving verbal feedback
* monitoring your progress

They are not responsible for:

* Providing real-time support over Slack. Slack is asynchronous communication.
* Solving problems for you

Joris Quist and Wouter Zorgdrager will be TAs for this year.

## Slide symbols

* **Q** A question with a known answer; this will be revealed, but we should
work together towards it!

* **D** A open discussion item; we need to think and discuss.

## Lecture notes

* Freely available on the web, on my homepage (http://gousios.org/teaching.html)

* You can print/download them before the lecture and bring them along to make
additional notes.

* I am looking forward to improve them! If you have suggestions etc, let me know!

```{r child="footer.Rmd", include=FALSE}
```
