---
title: "Algorithms and Data Structures: Things to remember"
author: "Georgios Gousios"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: revealjs::revealjs_presentation
bibliography: bibliography.bib
csl: ieee.csl
---

```{r child="header.Rmd", include=FALSE}
```

## Algorithms and complexity

- What is the divide and conquer technique?
- What is the $O$ and the $\Theta$ complexity?
- How can we calculate the complexity of recursive functions?
- How can we avoid the recomputation of recursive function calls?

## Basic data structures

- How do we implement a stack?
- How do we implement a linked list?
    - What types of lists exist?
- How do we implement a stack/queue using a linked list?
- What is the complexity of add/remove/search in all basic data structures?

## Trees

- What is the main benefit of trees?
- How are trees represented in memory? What types of nodes are there?
- What types of traversals can we do with trees?
- How do we build trees? What are the basic operations?
- What is a binary search tree? How can we implement one? What is the complexity
of searching in a BST?
- What are the use cases for more complex trees (AVL, RB and BTrees)? 
    - What are the complexities of basic ops?

## Graphs

- What is the relationship between trees and graphs?
- How do we represent graphs in memory?
- How do we search graphs? What is BFS and what is DFS?
- How can we calculate shortest paths?
- What is topological short and what are spanning trees?

## Sorting and searching

- What are the complexities of naive sorting algorithms?
- How can we implement quicksort recursively?
- How can we implement merge sort?
- What is sequential search and what is binary search?
- How can we implement binary search?

## Strings

- How are strings represented in memory? What is ASCII and what is Unicode?
- What is the intuition behind the Knuth-Morris-Pratt algorithm?
- How does string differencing work? 

## Optimization

- Why do we need optimization algorithms? What classes of problems do they solve?

```{r child="footer.Rmd", include=FALSE}
```
