---
title: "Trees"
date: "`r Sys.Date()`"
author: Joseph Hejderup
output: revealjs::revealjs_presentation
bibliography: bibliography.bib
csl: ieee.csl
---

```{r child="header.Rmd", include=FALSE}
```

# Tree as a data structure {data-background="figures/tree_american.jpg"}

## Why organize data as a tree structure?

- Humans like hierarchies and *hierarchical* ordering. Examples are *family
  tree*, *governance*, *species* 
- This form of systematic organization in a hierarchical order helps computers
  to localize and make decisions in a much more efficient way then linear data
  structures

## Example: Visiting a distant aunt

- You only meet her once every three years, she is from your dad's side. There
  is a huge family gathering next week and you need to find her address.
  Learning from the previous lecture, the data is organized in this fashion

![](./figures/tree_linear.svg)

## Example: Visiting a distant aunt

If we structure the family in a tree, we can find it her address in less time,
we know she is on the dad side, thus we avoid exploring half the family!

![](./figures/tree_nonlinear.svg)





## Definition of a tree

## Tree terminology

## Traversing a tree

# Binary Trees

## Properties

## Binary Search Trees (BST)

## Searching in a Binary Search Tree

## Tree Manipulation
- include cost of each operation

### Insertion
### Deletion

## Random Binary Tree

# Self-balancing binary search tree

## Treap

## AVL Tree

## WAVL Tree

## Scapegoat Tree

```{r child="footer.Rmd", include=FALSE}
```
