---
title: "Basic Data Structures"
author: "Maria Kechagia"
date: "18/10/2017"
output:
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Sets

In mathematics, a **set** is a collection of *distinct* elemets.

![](/Users/marki/Documents/myrepos/algo-ds/lectures/week2/figures/set-of-shapes.svg)

## Dynamic Sets

- Sets can be manipulated using appropriate algorithms. Then, sets can grow, shrink, or change over time.

- Dynamic sets can support basic operations such as:
*insert* elements to a set, *remove* elements from a set,
or *search* whether an element belongs to a set.

```script
  insert(S, x)
  # Add an element to a set S pointed to by x.
```

```script
  remove(S, x)
  # Given a pointer x to an element in a set S, remove x from S.
```

```script
  search(S, k)
  # Given a set S and a key k, return a pointer x to an element in S.
  # Output: x.key = k or NIL.
```

## Stacks and Queues (1)

- Stacks and queues are dynamic sets where the way an element can be removed from a set is prespecified.

- In a **stack**, the element to be removed is the one that was most recently inserted into the stack. We name this policy as: *last-in, first-out*, or *LIFO*.

- If a **stack** consists of elements ```S[1..S.top]```, ```S[1]``` is the element at the bottom of the stack and ```S[S.top]``` is the element at the top of the stack. If ```S.top = 0```, the stack is *empty*.

## Stacks and Queues (2)

- In a **queue**, the element to be removed is the one that belongs to the set for the longest time. We name this policy as: *first-in, first-out*, or *FIFO*.

- If a **queue** consists of elements ```Q[1..n]```, ```Q.heap``` points to its head and ```Q.tail``` points the *next* position where a new element can be added to the queue. When ```Q.heap = Q.tail```, it means that the queue is *empty*.

## Stack Operations

- The `insert` operation on a stack is called as `push`.
Additionally, the `remove` operation is called as `pop`.

- In the following example, we implement a stack as an *array* of *n* elements
to demonstate the `push` and `pop` operations.

```{r echo=FALSE}
  # The top element is the 4th one.
  x <- matrix(list(15,6,2,9,"N","N","N"), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]")))
  x
```

```{r echo=FALSE}
  # New elements (5th and 6th) are added to the end of the stack.
  x <- matrix(list(15,6,2,9,17,3,"N"), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]")))
  x
```

```{r echo=FALSE}
  # When the 6th element is removed, the next top element is the 5th
  x <- matrix(list(15,6,2,9,17,"[3]","N"), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]")))
  x
```

## Queue Operations

- The `insert` operation on a queue is called as `enqueue`.
Additionally, the `remove` operation is called as `denqueue`.

- In the following example, we implement a queue as an *array* of *n* elements
to demonstate the `enqueue` and `denqueue` operations.

```{r echo=FALSE}
  # Q.tail = 7 and Q.head = 12
  x <- matrix(list("N","N","N","N","N","N",15,6,9,8,4,"N"), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]","[8]","[9]","[10]","[11]","[12]")))
  x
```

```{r echo=FALSE}
  # Q.tail = 3 and Q.head = 7
  x <- matrix(list(3,5,"N","N","N","N",15,6,9,8,4,17), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]","[8]","[9]","[10]","[11]","[12]")))
  x
```

```{r echo=FALSE}
  # Q.tail = 3 and Q.head = 8
  x <- matrix(list(3,5,"N","N","N","N","[15]",6,9,8,4,17), nrow = 1, dimnames = list(c("S"), c("[1]","[2]","[3]","[4]","[5]","[6]","[7]","[8]","[9]","[10]","[11]","[12]")))
  x
```

## Lists

A **list** is a data structure that can store
items (usually) of the same type and
in a specific order.

Typically, we can create a list
by writing the values (items) we want to store
in a comma-separated list
and into square brackets.

```python
   list = [1, 2, 3, 4, 5]
   print(list)
```
  1, 2, 3, 4, 5

## Linked Lists

- A **linked list** is a data structure where objects
are arranged in a linear order and this order is determined
by poiters attached to each element of the list.

- There are two types of linked lists: *singly* and *doubled*.

- Linked lists support *all* the operations of dynamic sets.

- A **doubly linked list** ```L``` represents a dynamic set ```{1, 4, 9, 16}```,
where the attribute ```L.heap``` points to the head of the list
and ```L.tail``` points the last element of the list
(where after this there is no *next* element).
If ```L.heap = NIL```, it means that the list is *empty*.

- Given that ```x``` is an element of a list ```L```, 
```x.next``` points to the successor element of ```x```
and ```x.prev``` points to the predecessor element of ```x```.

## More Forms of Linked Lists

- If a list is **singly linked**, the ```prev``` pointer is omitted.

- If a list is **ordered**, the elements of the list
are arranged in linear order.
Then, the minimum element is the *head* of the list,
whereas the maxinum element is the *tail* of the list.

- If a list is **cincular**,
the ```prev``` pointer of the head of the list points to its tail,
and the ```next``` pointer of the tail of the list points to its head.

## Data Structure Algorithms

- Searching

- Ordering

- Hashing
